Climate indicators

### 2023-02-27 DRAFT_climate_indicators.Rmd
Fix conceptual workflow

And time is four dates covering four months in 1970:

@@

### Workflow

The general workflow is like this:

@@

    -   Import correct .nc file (loop though year 1961-1990) and subset to the correct attribute (above)

    -   Filter data by dates (optional)

    -   Aggregate across time within a year

    -   Aggregate across years to get median and sd values

    -   Intersect with accounting area polygons

2.  Calculate variable values

    -   Import correct .nc file (loop though the years 1961 to present) and subset to the correct attribute (above)

    -   Repeat the steps from when calculating the reference value, except for calculating the sd.

    -   Add layers to data cube already containing reference values

3.  Normalize climate variable at the individual grid cell level using the three reference values

4.  Write to file (for use with other ecosystem types)

5.  Mask by ecosystem type

@@

    -   Intersect with accounting area polygons

@@

### Calculate reference values

#### Import

I need to code a for loop here for importing the .nc files.

But for then selecting the attribute, simply do like this:

@@

Let's try it on the dummy data. Remember this data had three time steps.

@@

Our data has 365 time steps for each year.

@@

```{r}
start_month <-  "06"
end_month <- "08"
year_temp <- year(st_get_dimension_values(x, "time")[1])
start <- ym(paste(year_temp, start_month, sep="-"))
end <-  ym(paste(year_temp, end_month, sep="-"))
summer <- interval(start, end)
```

@@

```{r}
x_aug <- x_sub %>%
  filter(time %within% summer)
st_get_dimension_values(x_aug, "time")
```

Let me introduce some differences to Attribute_A between these two dates.

```{r}
x_aug$Attribute_A[,,,2] <- x_aug$Attribute_A[,,,1]*4
```

@@

For this example I want to calculate the mean temperature over the summer. I therefore need to aggregate over time. There are two ways.

*Method 1:*

@@

microOut <- microbenchmark(
st_apply =  x_aug %>%
  filter(band ==1) %>% # for our real data we don't need this filter
  st_apply(1:2, mean) %>%
  setNames(paste0("v_",temp_names)),
aggregate = x_aug %>%
  aggregate(by = "year", mean),
times=30
)
```

The time (and band) dimension is now gone as we have aggregated across it.

@@

`st_apply` is considerably faster, even wit the filtering step, but `aggregate` has the advantage of that it retains the time dimension, whereas for `st_apply` I need to set the attribute name to be the year. I will try to use `st_apply`.

@@

First, let me cut the _band_ dimension.
```{r}
#temp <- x_summerMean %>%
#  filter(band==1) %>%
#  adrop(drop=4) %>%
#  aperm(c(2,3,1))
#dim(temp)
```

```{r}
#oldTime <- st_get_dimension_values(x_summerMean_v2, "time")
#
## Function to add random noise
#myRandom <- function(x) x*rnorm(1,1,.05)
#
#y1 <- temp
#y2 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-1)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
#y3 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-2)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
#y4 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-3)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
#y5 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-4)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
#
#
## We combine the data into one cube:
#y <- c(y1, y2, y3, y4, y5)
```

>>>>

I will remove the _band_ dimension.
```{r}
x <- x %>% filter(band==1) %>% adrop()
```

Time is four dates covering four months in 1970:

```{r}
(oldTime <- st_get_dimension_values(x, "time"))
```

I'll add some extra seasonal variation between to the mix
```{r}
x$Attribute_A[,,2] <- x$Attribute_A[,,1]*1.2
x$Attribute_A[,,3] <- x$Attribute_A[,,1]*1.4
x$Attribute_A[,,4] <- x$Attribute_A[,,1]*1.1
```

Now I create another four copies of this data, adding some random noise and a continuous decreasing trend.

```{r}
y1 <- x
y2 <- x %>% st_apply(1:3, myRandom) %>% st_set_dimensions("time", values = oldTime %m+% years(-1)) %>% mutate(Attribute_A = Attribute_A*0.95)
y3 <- x %>% st_apply(1:3, myRandom) %>% st_set_dimensions("time", values = oldTime %m+% years(-2)) %>% mutate(Attribute_A = Attribute_A*0.90)
y4 <- x %>% st_apply(1:3, myRandom) %>% st_set_dimensions("time", values = oldTime %m+% years(-3)) %>% mutate(Attribute_A = Attribute_A*0.85)
y5 <- x %>% st_apply(1:3, myRandom) %>% st_set_dimensions("time", values = oldTime %m+% years(-4)) %>% mutate(Attribute_A = Attribute_A*0.80)

```{r}
# We combine the data into one cube for plotting:
temp <- c(y1, y2, y3, y4, y5)

ggplot() + 
  geom_stars(data = temp) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))+
  facet_wrap(~time, ncol=4)
```

Saving these to file to demonstrate the importing step of the workflow.

```{r, eval=F}
path <- "P:/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data"
saveRDS(y1, paste0(path, "/y1.rds"))
saveRDS(y2, paste0(path, "/y2.rds"))
saveRDS(y3, paste0(path, "/y3.rds"))
saveRDS(y4, paste0(path, "/y4.rds"))
saveRDS(y5, paste0(path, "/y5.rds"))
```

##### Dummy regions and ET map

Here is some dummy data for accounting areas (regions) and ecosystem occurrences as well.

```{r}
dummy_aa <- x %>%
  filter(time == oldTime[1]) %>%
  adrop() %>%
  mutate(accuntingAreas = rep(c(1,2), each=length(Attribute_A)/2)) %>%
  select(accuntingAreas)

tm_shape(dummy_aa) +
  tm_raster(style="cat")
```

```{r}
dummy_et <- x %>%
  filter(time == oldTime[1]) %>%
  adrop() %>%
  mutate(ecosystemType = rep(c(1,NA), length.out = length(Attribute_A))) %>%
  select(ecosystemType)

tm_shape(dummy_et) +
  tm_raster(style="cat")
```

@@

### Conseptual workflow

The general, conceptual workflow is like this:

1. Collate variable data series

    -   Import .nc files (loop though year 1961-1990) and subset to the correct attribute (above)

    -   Filter data by dates (optional) (`dplyr::filter`)

    -   Aggregate across time within a year (`stars::st_apply`)

    -   Join all data into one data cube (´stars:c´)

@@

    -   Aggregate across reference years to get median and sd values `st_apply`

    -   Join with existing data cube

2.  Calculate indicator values

    -   Normalize climate variable at the individual grid cell level using the three reference values `mutate %>% across`

4.  Write to file

5.  Mask by ecosystem type (this could also be done in step one to save time)

@@

    -   Intersect with accounting area polygons `sf:st_intersects`

@@

### Collate variable data series

I need to code a for loop for importing the .nc files.

For then selecting the attribute, simply do like this:

```{r}
# list attribute names
names(x)
```

@@

Let's try it on the dummy data. Remember this data had four time steps.

@@

Our real data has 365 time steps for each year.

@@

```{r}
start_month <-  "06"
end_month <- "08"
```

Then for each iteration I need to get the year as well
```{r}
(year_temp <- year(st_get_dimension_values(x, "time")[1]))
```

Then I can create a time interval object
```{r}
start <- ym(paste(year_temp, start_month, sep="-"))
end <-  ym(paste(year_temp, end_month, sep="-"))
(myInterval <- interval(start, end))
```

@@

```{r}
x_aug <- x_sub %>%
  filter(time %within% myInterval)
st_get_dimension_values(x_aug, "time")
```

@@

For this example I want to calculate the mean temperature over the summer. I therefore need to aggregate over time. There are two ways, either using `st_apply` or using `aggregate`.

@@

microOut <- microbenchmark(
st_apply =  x_aug %>%
  st_apply(1:2, mean) %>%
  setNames(paste0("v_",temp_names)),
aggregate = x_aug %>%
  aggregate(by = "year", mean),
times=30
)
```

The time dimension is now gone as we have aggregated across it.

@@

`st_apply` is slightly faster, but `aggregate` has the advantage of that it retains the time dimension, whereas for `st_apply` I need to set the attribute name to be the year. I will try to use `st_apply`.

But let me see if I can save time by masking to ET before doing this operation.

```{r, fig.cap="Demonstrating the effecty of maskig the dummy data using a perfectly aligne raster ET map."}
x_masked <- x_aug
x_masked[is.na(dummy_et)] <- NA
plot(x_masked[,,,1])
```

```{r, fig.cap="There is no performance increas by making before aggregating"}
autoplot(
  microbenchmark(
No_masking =  x_aug %>%
  st_apply(1:2, mean) %>%
  setNames(paste0("v_",temp_names)),
Masking =  x_masked %>%
  st_apply(1:2, mean) %>%
  setNames(paste0("v_",temp_names))
 ), times=30
)

@@
